---
title: "Spanish Stress/Tense F0 Study — Model Specifications & Results"
subtitle: "First-three-words acoustic analysis (article / noun / verb); present–paroxytone vs past–oxytone"
author: "Kaylee Fernandez"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

# Overview

This document records every statistical model fit in the analysis, with its
**dependent variable**, **independent (fixed) variables**, **random
intercepts/slopes**, and **result**. Two families of model were run:

1. **Whole-utterance contour GAMMs** (F0, intensity) over a per-syllable time
   index spanning Article → Noun → Verb (`analyze_3words_F0.R`).
2. **N-s2 windowed tests** isolating the noun's second syllable, asking whether
   it carries an *anticipatory* cue to the upcoming verb's stress
   (`test_Ns2_cues.R`).

Design constants throughout:

| Element | Value |
|---|---|
| Participants | 5 (each produced both conditions) |
| Items | 90 two-syllable-noun items (90 of 90 kept) |
| Condition `a` | Present tense → **paroxytone** (penult stress, e.g. BUS-ca) |
| Condition `b` | Past tense → **oxytone** (final stress, e.g. bus-CO) |
| Sampling | 10 points per syllable; 5 syllables (Art, N-s1, N-s2, V-s1, V-s2) |
| Rows extracted | 90,000 total; 74,070 voiced (82.3%) for F0 |
| Effect direction | All "b − a" = past − present |

> **Sample-size caveat.** With 5 participants the by-participant random-smooth
> structure is at the minimum the design supports. Results are reported as
> inferential (the script did **not** enter pilot/descriptive mode), but the
> participant-level estimates are thinly supported and should be interpreted
> with that in mind.

---

# Part 1 — Whole-utterance contour GAMMs

Both contour models share the same specification, fit with
`mgcv::bam(method = "fREML", discrete = TRUE)`, following Wieling (2018) and
Sóskuthy (2021). The contrast is carried by an **ordered factor** `conditionO`
(present = reference), giving a parametric **height** term and a smooth
**shape** (difference-smooth) term. A companion **binary-smooth** model
provides a single combined "any difference" test. An AR(1) error model
corrected residual autocorrelation; the factor-smooth basis (tp vs cr) was
chosen by AIC.

## Common structure

| Component | Specification |
|---|---|
| **Independent (fixed)** | `conditionO` (ordered: present vs past) |
| **Reference smooth** | `s(time_bin, k = 25)` — population contour |
| **Difference smooth** | `s(time_bin, by = conditionO, k = 25)` — shape difference |
| **Random intercept (item)** | `s(item_id, bs = "re")` |
| **Random reference smooth (participant)** | `s(time_bin, participant, bs = "fs", m = 1)` |
| **Random difference smooth (participant)** | `s(time_bin, participant, by = conditionO, bs = "fs", m = 1)` |
| **Time axis** | per-syllable index 0–5 (Art 0–1, N-s1 1–2, N-s2 2–3, V-s1 3–4, V-s2 4–5) |

The by-participant **reference + difference** factor smooths implement
Sóskuthy's (2021) within-item recommendation: the difference smooth shrinks the
by-participant contrast (analogous to `(1 + condition | participant)`),
restoring power while keeping Type I error nominal.

```{r model-formula-contour}
# Schematic (same for both measures; yvar = "f0_st" or "intensity_db")
yvar ~ conditionO +
  s(time_bin, k = 25) +
  s(time_bin, by = conditionO, k = 25) +
  s(item_id, bs = "re") +
  s(time_bin, participant, bs = "fs", m = 1, xt = <basis>, k = 25) +
  s(time_bin, participant, by = conditionO, bs = "fs", m = 1, xt = <basis>, k = 25)
```

## Model 1 — F0 contour

| Field | Value |
|---|---|
| **Dependent variable** | F0 in semitones re 100 Hz (`f0_st`) |
| **Family** | Scaled-t (`scat`; df = 3, σ = 0.953) — heavy-tail robust |
| **Factor-smooth basis** | cr (selected by AIC; cr 270213.7 vs tp 270234.6) |
| **AR(1) rho** | −0.038 |
| **N** | 74,070 (voiced points) |
| **Fit** | R²(adj) = 0.778; deviance explained = 74% |

**Results (summary of smooth/parametric terms):**

| Term | Estimate / F | edf | p |
|---|---|---|---|
| Intercept | 10.15 st | — | < .001 |
| `conditionOb` (height: past − present) | **−0.23 st**, SE 0.05, t = −4.36 | — | **< .001** |
| `s(time_bin)` (reference contour) | F = 2.48 | 17.20 | < .001 |
| `s(time_bin):conditionOb` (shape diff.) | F = 3.46 | 8.00 | **< .001** |
| `s(item_id)` (item RE) | F = 20.29 | 84.84 | < .001 |
| `s(time_bin,participant)` (ref. fs) | F = 4761.83 | 96.93 | < .001 |
| `s(time_bin,participant):conditionOb` (diff. fs) | F = 17.43 | 73.32 | < .001 |
| **Combined (binary smooth)** | F = 4.98 | 9.00 | **< .001** |

- **Significant present/past F0 difference** in both height and shape; survives
  Bonferroni (.025) on the split tests.
- Difference is **localized to the verb**: the pointwise CI excluded zero across
  **V-s1 (72%) → V-s2 (24%)**, mean Δ = 0.72 st. Article and noun regions: no
  difference.
- *k* diagnostics clean (edf well below k′; k-index p > .15 for all smooths).

## Model 2 — Intensity contour

| Field | Value |
|---|---|
| **Dependent variable** | Intensity in dB (`intensity_db`) |
| **Family** | Gaussian |
| **Factor-smooth basis** | cr (AIC: cr 545319.7 vs tp 545321.2) |
| **AR(1) rho** | 0.394 |
| **N** | 89,319 |
| **Fit** | R²(adj) = 0.449; deviance explained = 45% |

**Results:**

| Term | Estimate / F | edf | p |
|---|---|---|---|
| Intercept | 56.48 dB | — | < .001 |
| `conditionOb` (height: past − present) | **−0.20 dB**, SE 0.08, t = −2.58 | — | **.010** |
| `s(time_bin)` (reference contour) | F = 53.17 | 23.54 | < .001 |
| `s(time_bin):conditionOb` (shape diff.) | F = 5.26 | 12.69 | **< .001** |
| `s(item_id)` (item RE) | F = 26.59 | 85.81 | < .001 |
| `s(time_bin,participant)` (ref. fs) | F = 23.61 | 78.41 | < .001 |
| `s(time_bin,participant):conditionOb` (diff. fs) | F = 0.20 | 11.91 | .005 |
| **Combined (binary smooth)** | F = 5.05 | 13.69 | **< .001** |

- **Significant present/past intensity difference** in height and shape.
- Difference window: **V-s1 (50%) → V-s2 (27%)**, mean Δ = 1.16 dB — again
  verb-localized.
- ⚠ **k flag:** `s(time_bin)` reference smooth had edf 23.5 / k′ 24 (k-index
  0.97, p = .04). The reference smooth is near its basis ceiling. *Recommended
  robustness check: refit with `k = 35` and confirm the difference smooth and
  its p-value are stable.*

---

# Part 2 — N-s2 windowed tests (anticipatory cue)

These isolate the **noun's second syllable (N-s2)** and ask whether it
distinguishes the upcoming verb's stress *before the verb begins*. Duration,
intensity, and mean F0 were tested with linear mixed-effects models
(`lme4`/`lmerTest`); F0 contour shape with a GAMM difference smooth restricted
to N-s2 points. N-s2 tokens = 1,800 (900 present, 900 past).

## Common LMM structure (duration, intensity, mean F0)

| Component | Specification |
|---|---|
| **Independent (fixed)** | `condition` (present vs past) |
| **Random intercept** | `(1 | participant)`, `(1 | item_id)` |
| **Random slope** | `(condition | participant)` — *attempted* |
| **Estimation** | REML, `bobyqa`; auto-fallback to intercepts-only if singular |

```{r model-formula-lmm}
# Maximal (attempted):
dv ~ condition + (1 + condition | participant) + (1 | item_id)
# Fallback if singular:
dv ~ condition + (1 | participant) + (1 | item_id)
```

## Model 3 — N-s2 Duration (LMM)

| Field | Value |
|---|---|
| **Dependent variable** | N-s2 duration (ms) |
| **Fixed** | `condition` |
| **Random structure used** | **Intercepts-only** — maximal model was singular |
| **Result** | b (past − present) = **−9.51 ms**, SE 1.35, t = **−7.02**, **p < .0001** |

→ **Robust anticipatory cue.** N-s2 is reliably ~9.5 ms *shorter* before an
oxytone (past) verb. Survives Bonferroni comfortably. *(Present 157 ms vs past
148 ms descriptively.)*

## Model 4 — N-s2 Intensity (LMM)

| Field | Value |
|---|---|
| **Dependent variable** | N-s2 mean intensity (dB) |
| **Fixed** | `condition` |
| **Random structure used** | Intercepts-only (maximal singular) |
| **Result** | b = +0.02 dB, SE 0.14, t = 0.13, p = .898 |

→ **Null.** No anticipatory intensity cue.

## Model 5 — N-s2 Mean F0 / height (LMM)

| Field | Value |
|---|---|
| **Dependent variable** | N-s2 mean F0 (semitones) |
| **Fixed** | `condition` |
| **Random structure used** | **Maximal** `(1 + condition | participant) + (1 | item_id)` (converged) |
| **Result** | b = +0.21 st, SE 0.21, t = 0.99, p = .378 |

→ **Null** for mean F0 (height). *(Note: this is the one N-s2 model where the
full random-slope structure held.)*

## Model 6 — N-s2 F0 contour shape (GAMM)

Tests whether the F0 *trajectory through* N-s2 differs by condition even when
the mean does not. Time axis = within-syllable progress (0 → 1).

| Component | Specification |
|---|---|
| **Dependent variable** | `f0_st` (voiced N-s2 points) |
| **Independent (fixed)** | `conditionO` (ordered) + `s(t, k = 8)` reference + `s(t, by = conditionO, k = 8)` difference |
| **Random intercept (item)** | `s(item_id, bs = "re")` |
| **Random reference smooth (participant)** | `s(t, participant, bs = "fs", m = 1, k = 8)` |
| **Random difference smooth (participant)** | `s(t, participant, by = conditionO, bs = "fs", m = 1, k = 8)` |

```{r model-formula-ns2shape}
f0_st ~ conditionO +
  s(t, k = 8) +
  s(t, by = conditionO, k = 8) +
  s(item_id, bs = "re") +
  s(t, participant, bs = "fs", m = 1, k = 8) +
  s(t, participant, by = conditionO, bs = "fs", m = 1, k = 8)
```

**Results:**

| Term | Estimate / F | edf | p |
|---|---|---|---|
| Shape difference `s(t):conditionOb` | F = **6.41** | 1.00 | **.011** |
| Height offset `conditionOb` (parametric) | +0.19 st | — | .343 |

→ **Marginal / suggestive.** The within-syllable F0 *tilt* differs by condition
(edf ≈ 1 → near-linear), although mean F0 does not. At p = .011 it just clears
α = .05 but is **borderline under Bonferroni** (.0125 for 4 tests) and visually
the predicted contours overlap heavily. *Recommended: refit with `k = 10` to
confirm robustness; report as suggestive, not confirmatory.*

---

# N-s2 summary table

| Cue | Effect (past − present) | Statistic | p | Bonferroni (α ≈ .0125) |
|---|---|---|---|---|
| **Duration** | **−9.51 ms** | t = −7.02 | **< .001** | ✅ survives |
| Intensity | +0.02 dB | t = 0.13 | .898 | — null |
| F0 mean (height) | +0.21 st | t = 0.99 | .378 | — null |
| F0 shape (tilt) | (shape) | F = 6.41 | .011 | ⚠ marginal |

**Take-away.** The noun's final syllable carries a **robust durational**
anticipatory cue to the upcoming verb's stress, plus a **subtle (marginal) F0
contour-tilt** cue; there is **no** anticipatory cue in intensity or in mean F0.
Because mean F0 is null while shape is (marginally) significant, the contour
approach is doing work a syllable-mean analysis would miss — a methodological
point in favour of the GAMM, even if the substantive F0 claim stays cautious.

---

# Variable glossary

| Variable | Role | Levels / unit |
|---|---|---|
| `f0_st` | DV | F0, semitones re 100 Hz |
| `intensity_db` | DV | Intensity, dB |
| `syll_dur_ms` / `dur_ms` | DV | Syllable duration, ms |
| `condition` / `conditionO` | Fixed IV | a = present/paroxytone; b = past/oxytone |
| `time_bin` | Smooth covariate | per-syllable index 0–5 (whole-utterance models) |
| `t` (`within_syll`) | Smooth covariate | within-syllable progress 0–1 (N-s2 shape model) |
| `participant` | Random grouping | 5 speakers |
| `item_id` | Random grouping | 90 items |

# Reproducibility notes

- Pipeline: `align_cut_audios.py --all` (MFA → 4-tier TextGrids) →
  `extract_f0_3words.praat` (90,000-row table) → `analyze_3words_F0.R`
  (Models 1–2) → `test_Ns2_cues.R` (Models 3–6).
- Outputs: `plots/`, `tables/` (incl. `APA_results.txt`,
  `Ns2_test_results.csv`), `models/f0_intensity_models.RData`.
- Two flagged robustness checks: (i) intensity Model 2 with `k = 35`;
  (ii) N-s2 shape Model 6 with `k = 10`.
- All models estimated in R with `mgcv` (GAMMs) and `lme4`/`lmerTest` (LMMs).
